---
title: "Cycling Hour Records"
date: "2023-07-07"
description: "Time trials and tribulations."
execute: 
   warning: false
   message: false
filters:
   - lightbox
lightbox: auto
draft: true
---

```{r}
#| label: libraries
#| include: false

library(here)
library(tidyverse)
library(kableExtra)
library(scales)
library(viridis)
library(ggrepel)
```

## Introduction

One of the most prestigious achievements in professional cycling is the hour record: travel farther on a bicycle in one hour than anyone has before.

## History

Henri Desgrange, the bicycle racer and sports journalist who organized the Tour de France in its earliest years, set the first official hour record at 35.325 km on the Vélodrome Buffalo in Paris in 1893. Two months later, a Mlle^[Her first name seems to be lost to history.] de Saint-Saveur set the women's record at 26.012 km on the same track.

Eight decades later, Eddy Merckx would break the men's record by riding 49.431 km. Merckx was by that time undisputedly the greatest cylist of all time, but despite taking advantage of the lower aerodynamic drag offered by Mexico City's high altitude, he described it as the hardest ride of his life. Maria Cressari broke the women's record with a 41.471 km ride on the same track a month later.

The 1980s and 90s were a period of controversy for the hour record. A series of improvements on the records of Merckx and Cressari were achieved, but many attributed these to more aerodynamic equipment and riding postures. The battle over the hour record between the unconventional cyclist Graeme Obree and the UCI^[Professional cycling's regulatory body.] was dramatized in the 2006 film [The Flying Scotsman](https://www.imdb.com/title/tt0472268/). The ultimate result was a decision split the hour record into two categories: one requiring the use of equipment similar to what was available to Merckx and Cressari in the early 70s, and another allowing basically anything that was arguably a bicycle. This change was retroactive, so the official men's hour record was reverted to Merckx's. 

While Cressari's record had been improved upon five times by the early 2000s, the men's record was just barely broken^[By a margin of only 10 m!] by Chris Boardman and then by Ondřej Sosenka. The former had the advantage of shoes which locked into the pedals and the latter was probably doping at the time he the broke record. There followed a decade of disinterest in the hour record.

The UCI revived the hour record in 2014 by changing the rules again. Going forward, the hour record would use the same equipment as Olympic track cycling races. The result was a burst of new hour record attempts including several successes. The current record holders are [Filippo Ganna](https://www.procyclingstats.com/rider/filippo-ganna/) and [Ellen van Dijk](https://www.procyclingstats.com/rider/ellen-van-dijk) with 56.792 km and 49.254 km respectively.

## Data

We can classify successful record attempts into four types: **historical** records set before Merckx's, **classic** records set under the restrictive equipment rules, **absolute** records set under the permissive rules, and **unified** records set since the unification of the equipment rules with track racing.

```{r}
#| label: data-preparation
#| message: false

read_csv("hour_records.csv", col_types = "ccffn") |>
  mutate(date = mdy(date)) ->
  hour_records

hour_records |>
  kable(caption = "Cycling Hour Records") |>
  scroll_box(height = "5in")
```

## Visualization

Expand the code for the second plot to see a useful trick that allows one to modify the data set underlying a `ggplot` object, in this case to filter out the historical records.

::: panel-tabset

### 1893-present 

```{r}
#| label: chart-1
#| warning: false
#| code-fold: true

hour_records |>
  ggplot() +
  # aesthetic mapping
  aes(x = date, y = distance, colour = type, fill = type, label = name) +
  # visual elements representing the data
  geom_line(colour = "#b4b4b4") +
  geom_point(size = 2) +
  # scales
  scale_y_continuous(breaks = seq(25,60,5), minor_breaks = seq(25, 60, 1)) +
  scale_x_date(breaks = as.Date(paste0(seq(1890, 2030, by = 5), "-01-01")),
               labels = date_format("%Y"),
               guide = guide_axis(n.dodge = 2),
               limits = as.Date(c("1890-01-01", "2025-01-01")),
               expand = c(0, 0)) +
  scale_colour_viridis(discrete = TRUE) +
  scale_fill_viridis(discrete = TRUE) +
  # facets
  facet_wrap(facets = vars(gender), 
             scales = "free_y", 
             nrow = 2, 
             strip.position="right") +
  # labels
  labs(title = "Cycling Hour Records", 
       y = "Distance (km)", 
       x = "Date") +
  # theming
  theme_bw() +
  theme(legend.position = "none",
        legend.title = element_blank(),
        panel.grid.minor.x = element_blank()) ->
  p

p
```

### 1972-present

```{r}
#| label: chart-2
#| warning: false
#| code-fold: true

# change the data embedded in the previous plot
p$data |>
  filter(type != "Historical") ->
  p$data

# re-plot it with text labels and new scales
p +
  geom_text_repel(size = 2, colour = "black") +
  scale_x_date(breaks = as.Date(paste0(seq(1970, 2030, by = 5), "-01-01")),
               labels = date_format("%Y"),
               limits = as.Date(c("1970-01-01", "2025-01-01")),
               expand = c(0, 0)) +
  scale_y_continuous() +
  scale_colour_viridis(discrete = TRUE, begin = 0.33) +
  labs(title = "Cycling Hour Records, 1972-present")
```
:::

## Observations

These charts highlight how little progress was made on the men's record during the classic era. We can also see that the unified records have overtaken the old "anything goes" absolute records. Finally, it's interesting to note that the current women's record is just slightly shy of Merckx's.

## References

- [Hour record (Wikipedia)](https://en.wikipedia.org/wiki/Hour_record)
- [Eddy Merckx and the Hour Record](https://bikeraceinfo.com/riderhistories/Merckx-Hour-Record.html)
- [The Retrogrouch: The Hour Record ](http://bikeretrogrouch.blogspot.com/2014/05/the-hour-record.html)